From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-18 Pi+/paper/reversible2.tex
restructure and work on denot sem

\section{Reversible Programming Languages}~\label{sec:reversible}

The practice of programming languages is replete with \emph{ad hoc} instances of reversible computations: database
transactions, mechanisms for data provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional memories, continuations, backtracking
search, and multiple-level undo features in commercial applications. In the early nineties,
\citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of reversibility. But intensive
research in full-fledged reversible models of computations and reversible programming languages was only sparked by the
discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275, Hey:1999:FCE:304763,fredkin1982conservative}, and by the
potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started
with a conventional programming language, e.g., an extended
$\lambda$-calculus, and either:
\begin{enumerate}
\item extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
\item imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
\end{enumerate}
More modern approaches recognize that reversible programming languages require a fresh approach and should be designed
from first principles without the detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047,
  rc2011,James:2012:IE:2103656.2103667,Carette2016}.

%%%%%%%%%%%%%%%%%
\subsection{The $\Pi$ Family of Languages}

A natural candidate for a semantic foundation for reversible programming languages is the notion of type isomorphism
(equivalence). Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667} and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}.

Focusing on finite types, the building blocks of type theory are: the
empty type ($\mathbb{0}$), the unit type ($\mathbb{1}$) containing
just one value $\Acon{tt}$, the sum type ($+$) containing values of
the form $\inlv{v}$ and $\inrv{v}$, and the product ($\times$) type
containing pairs of values $(v_1,v_2)$. One may view each type $A$
as a collection of physical wires that can transmit $|A|$ distinct
values where $|A|$ is a natural number that indicates the size of a
type, computed as: $| \mathbb{0} | = 0$; $| \mathbb{1} | = 1$;
$| A + B | = | A | + |B |$; and
$| A \times B | = | A | * |B |$.  Thus the type
$\mathbb{B} = \mathbb{1} + \mathbb{1}$ corresponds to a wire that can
transmit one of two values, i.e., bits, with the convention that
$\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and
$\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The type
$\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
collection of wires that can transmit three bits. From that
perspective, a type isomorphism between types $A$ and $B$
(such that $|A|=|B|=n$) models a \emph{reversible}
combinational circuit that \emph{permutes} the $n$ different
values. These type isomorphisms are collected in Fig.~\ref{pi-terms}.
\begin{figure}[t]
{\scalebox{\scalef}{$%
%%\noindent\begin{minipage}{.7\linewidth}
\begin{array}{rrcll}
\idc :& A & \iso & A &: \idc \\
\\
\identlp :&  \mathbb{0} + A & \iso & A &: \identrp \\
\swapp :&  A + B & \iso & B + A &: \swapp \\
\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
\identlt :&  \mathbb{1} \times A & \iso & A &: \identrt \\
\swapt :&  A \times B & \iso & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
\absorbr :&~ \mathbb{0} \times A & \iso & \mathbb{0} ~ &: \factorzl \\
\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
  \end{array}$}}

{\scalebox{\scalef}{%
\Rule{}
{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
{\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
{}

\Rule{}
{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
{\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
{}

\Rule{}
{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
{\jdg{}{}{c_1 \otimes c_2 : A \times C \iso B \times D}}
{}
}}
\caption{$\Pi$-terms, combinators, and their types.}
\label{pi-terms}
\end{figure}

Each line in the top part of the figure introduces a pair of dual constants that witness the type isomorphism in the
middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence
relation by adding the three constructors at the bottom of the figure that witness equivalence and compatible
closure. Although austere, this combinator-based language has the advantage of being amenable to formal analysis for at
least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident connections to type
theory and category theory and hence possesses a rich algebraic structure which we exploit in the remainder of the
paper.

% Specifically, the type isomorphisms of~$\Pi$ are sound and complete for all permutations on finite
% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980}. Algebraically, these types and
% combinators form a \emph{commutative semiring} (up to type isomorphism). Logically, they form a superstructural logic
% capturing space-time trade-offs~\cite{superstructural}. Categorically, they form a \emph{distributive bimonoidal
%   category}~\cite{laplaza72}.

% For the fragment over finite types, the syntax of the language $\Pi$ consists of several sorts:

%  {\scalebox{\scalef}{$%
% \begin{array}{lrcl}
% \textit{Value types} & A,B,C,D &::=& \mathbb{0} \alt \mathbb{1} \alt A+B \alt A\times B \\
% \textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
% \textit{Program types} &&& A \leftrightarrow B \\
% \textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
% \end{array}$}}

\begin{figure}[t]
{\scalebox{\scalef}{%
$\begin{array}{rclcrcl}
\delta(\identlp,~\inrv{v}) & = & v & & \delta(\identrp,~v)        & = & \inrv{v}\\
\delta(\swapp,~\inlv{v})     & = & \inrv{v}\\
\delta(\swapp,~\inrv{v})     & = & \inlv{v}\\
\delta(\assoclp,~\inlv{v})          & = & \inlv{(\inlv{v})} & & \delta(\assocrp,~\inlv{(\inlv{v})}) & = & \inlv{v}         \\
\delta(\assoclp,~\inrv{(\inlv{v})}) & = & \inlv{(\inrv{v})} & & \delta(\assocrp,~\inlv{(\inrv{v})}) & = & \inrv{(\inlv{v})}\\
\delta(\assoclp,~\inrv{(\inrv{v})}) & = & \inrv{v}          & & \delta(\assocrp,~\inrv{v})          & = & \inrv{(\inrv{v})}\\
\delta(\identlt,~(\Acon{tt},v)) & = & v & & \delta(\identrt,~v)             & = & (\Acon{tt},v)\\
\delta(\swapt,~(x,y))         & = & (y,x)    \\
\delta(\assoclt,~(x,(y,z))) & = & ((x,y),z) & & \delta(\assocrt,~((x,y),z)) & = & (x,(y,z))\\
\delta(\dist,~(\inlv{x},z)) & = & \inlv{(x,z)} & & \delta(\factor,~\inlv{(x,z)}) & = & (\inlv{x},z)\\
\delta(\dist,~(\inrv{y},z)) & = & \inrv{(y,z)} & & \delta(\factor,~\inrv{(y,z)}) & = & (\inrv{y},z)
\end{array}$}}
\caption{Semantics of base combinators}\label{fig:delta}
\end{figure}

\begin{figure}[t]
  {\scalebox{\scalef}{%
 $\begin{array}{c}
 \infer{\interp{c}{v} \Downarrow \delta(c,v)}{c ~\mbox{base combinator}}
\qquad
\infer{\interp{c_1\oplus c_2}{\inlv{v}} \Downarrow \inlv{w}}
    {\interp{c_1}{v} \Downarrow w}
\qquad
\infer{\interp{c_1\oplus c_2}{\inrv{v}} \Downarrow \inrv{w}}
    {\interp{c_2}{v} \Downarrow w} \\
\\
 \infer{\interp{\idc}{v} \Downarrow v}{}
\qquad
\infer{\interp{c_1\fatsemi c_2}{v} \Downarrow v_2}
    {\interp{c_1}{v} \Downarrow v_1 \quad \interp{c_2}{v_1} \Downarrow v_2}
\qquad
\infer{\interp{c_1\otimes c_2}{(v_1 , v_2)} \Downarrow (w_1 , w_2)}
    {\interp{c_1}{v_1} \Downarrow w_1 \quad \interp{c_2}{v_2} \Downarrow w_2}
\end{array}$}}
\caption{$\Pi$-interpreter}\label{fig:pi:interp}
\end{figure}

\noindent To formalize the semantics of $\Pi$, we first specify the
semantics of the base combinators in Fig.~\ref{fig:delta} using a
function $\delta$. It is clear that this function has an inverse
$\delta^\dagger$ such that $\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter
in Fig.~\ref{fig:pi:interp}.

%%%%%%%%%
\subsection{Examples}
\label{sec:langRev-examples}
\label{examples}

\note{Revisit to pick examples relevant to the rest of the paper; also unify notation}

\paragraph*{Booleans}
Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
Boolean negation is straightforward to define:

\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}

\ensuremath{\mathit{not} = \swapp}

\noindent
It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
vice versa.

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}

\noindent We can express a 3-bit word reversal operation as follows:

\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this paper, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

\noindent On the first line is the initial value. On each subsequent
line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
results from applying this combinator to the value on the previous
line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
the bits in reverse order.

We can also draw out the graphical representation of the 3-bit reverse
combinator. In the graphical representation, it is clear that the
combinator achieves the required shuffling.

% \inkscape{reverse-3-bit.pdf}

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
then we essentially have an `if' expression.

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}

% \inkscape{if-c1-c2.pdf}

The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_1}
processed by the distribute operator \ensuremath{\dist}, which converts it into
a value of type \ensuremath{(1 {\prodtype}  b_1){\sumtype}(1 {\prodtype}  b_1)}. In the
\ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
the value of type~\ensuremath{b_1}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied
to get the final result of type \ensuremath{(1{\sumtype}1) {\prodtype} b_2}.

\paragraph*{Logic Gates}
There are several universal primitives for conventional (irreversible)
hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
of reversible hardware circuits, the canonical universal primitive is
the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
bit is negated. In a traditional language, the Toffoli gate would be
most conveniently expressed as a conditional expression like:

\noindent
\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}

We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
of the conditional derived above. If the \ensuremath{\mathit{bool}} is
\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
% \includegraphics{inkscape/cnot.pdf}
}
\end{center}

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

\begin{center}
\scalebox{1.6}{
% \includegraphics{inkscape/toffoli.pdf}
}
\end{center}

%%%%%%%%%
\subsection{A Language of Equivalences between Type Equivalences}
\label{langeqeq}

>>>>

\section{A Reversible Programming Language: Groupoid Semantics}
\label{sec:reversibletwo}
\label{langeqeq}

### 2021-06-18 Pi+/paper/ufin.tex
restructure and work on denot sem

In order to defiine the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified. \note{Is this the correct informal explanation of Ufin?}

>>>>

Previous semantics adequate for reasoning about evaluation and for checking the equivalence of two circuits.

Want NBE, want rewriting system of program equivalence, what else? So we need to move to proof-relevant higher-category
groupoid semantics instead of set-theoretic.

The lambda calculus is classically presented with types, terms, and
$\beta\eta$-equations. This can be turned into a 1-category, with 0-cells for
types, and 1-cells for programs quotiented by congruence closure of the
$\beta\eta$-equations.

Then there is the free CCC, which is a 1-category constructed using
0-cells and 1-cells with universal properties that hold upto the notion
of equality that from the ambient metatheory.

Lambek/Scott show that these are equivalent. In modern language, this is
a 1-equivalence of (1,1)-categories.

Now, lets work without quotients in a proof-relevant metatheory.

The lambda calculus can now be presented with 0-cells for types, 1-cells
for terms, and 2-cells for equations/rewrites. This is a
(2,1)-category. Then, you can give a description of the free 2-CCC,
where universal properties hold upto canonical isomorphism.

Then we can show that there is a 2-equivalence of these
(2,1)-categories. (Actually, it can be weaker than (2,1)-categories, in
fact it holds for bicategories.)

Now, lets go to our result.

For Pi/Pi+, we present it as a (2,0)-category (or 1-groupoid à la HoTT),
with 0-cells for types, 1-cells for reversible programs, 2-cells for
rewrites/equations. And we present different descriptions (Pi+, Ufin)
of the free (2,0)-SMC (or 1-SMG) and show they're 2-equivalent.

So this is just a difference of presentation, in going from the (1,r) to
the (2,r) category. We have simply made the equations/rewrites
first-class in the presentation.

Now the question is, what does it mean for a compiler?

My opinion is that it is still the same, what rewrites to apply depends
on the choices made by the compiler writer, but we can think about that.

%%%
\subsection{Univalent Subuniverses}
\label{sec:univalent}

In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified. \note{Is this the correct informal explanation of Ufin?}

 In this section, we introduce some basic concepts and notation that we use in Homotopy Type Theory. Then, we define
univalent subuniverses and discuss some specific examples.

\subsection{The Type Theory}~\label{subsec:type-theory}

We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

\note{It is important that we work in HoTT using it as a metatheory, we care about proof relevance, constructivity, and
  working with higher-dimensional algebraic structures, such as groupoids. Our results are also formalised using general
  tools and principles from HoTT, and computer checked using a proof assistant, using Agda and the HoTT-Agda library.}

\subsubsection{Identity Types}

\vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}

In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.

\todo{Say why it is a groupoid by listing some groupoid laws.}

Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
by

\[
  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
\]

Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have

\[
  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\]

\todo{Why is the topological viewpoint important, or saying fibrations? Explain the intuition using classifying spaces,
  maybe draw some diagrams.}

From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ to
the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have

\[
  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
\]

where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.

Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define

\[
  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

\todo{Explain motivation.}

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
\end{definition}

\subsubsection{Univalence}

Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
the following function is an equivalence.

\[
  \ua : A \id_{\UU} B \to A \eqv B
\]

\todo{Revise.}

Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.

\subsubsection{Higher Inductive Types}

\todo{Explain homotopy types, $\hProp$, $\hSet$, etc.}

\vc{These are placeholder definitions, need informal explanations and references to the book.}

\begin{definition}[Propositional Truncation]
  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
  generated by the following constructors,
  \begin{itemize}
    \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
    \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
  \end{itemize}
  such that, given any type $B$ with
  \begin{itemize}
    \item a function $g : A \to B$,
    \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
  \end{itemize}
  there is a unique function $f : \Trunc{A} \to B$ such that,
  \begin{itemize}
    \item $f(\trunc{a}) \equiv g(a)$
    \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
  \end{itemize}
\end{definition}

\begin{definition}[Set Quotient]
  Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
  higher inductive type generated by
  \begin{itemize}
    \item an inclusion function $q : A \to \quot{A}{R}$,
    \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
    \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
  \end{itemize}
  with \review{an appropriate induction principle.}
\end{definition}

We recall that quotients in HoTT are \emph{effective}, that is, if $R$ is an equivalence relation, we have
$R(x,y) \eqv (q(x) \id_{\quot{A}{R}} q(y))$.

\subsection{Univalent Subuniverses}

\todo{Explain motivation. Starting from a univalent universe which classifies all types, we want to define a subuniverse
  which classifies only certain types, for example, types that satisfy some property that we want.}

\begin{definition}[Universe]
  A universe \`{a} la Tarski is given by the following pieces of data,
  \begin{itemize}
    \item a code $U : \UU$,
    \item a decoding function $\El : U \to \UU$.
  \end{itemize}
  If $\El$ is a univalent fibration, $U$ is a univalent universe.
\end{definition}

\begin{definition}[Subuniverse]
  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $\forall x, \isProp{P(x)}$. A
  subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
\end{definition}

\begin{proposition}[Univalent Subuniverse]
  Subuniverses generated by subtypes are univalent.
\end{proposition}

\begin{proof}
  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{proof}

\begin{example}[$\BAut$]
  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
  for the image of the inclusion of $T$.
\end{example}

\begin{example}
  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
\end{example}

\begin{definition}[$\Fin$]
  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
  two different ways,
  \begin{gather*}
    \begin{aligned}
      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
    \end{aligned}
    \begin{aligned}
      & \Fin[0] \defeq \bot \\
      & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
    \end{aligned}
  \end{gather*}
  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
\end{definition}

\begin{example}
  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
  image of the inclusion of $\Fin[n]$.
\end{example}

\begin{definition}[$\isFin$]
  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
  \[
    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
  \]
  Note that the natural number $n$ need not be truncated, as justified below.
\end{definition}

\begin{proposition}
  For any type $X$, $\isFin[X]$ is a proposition.
\end{proposition}

\begin{proof}
  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
  applying the first projection.
\end{proof}

\begin{example}
  The univalent subuniverse of \emph{all finite types} is given by
  \[
    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
  \]
  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
  denotational semantics to interpret our reversible programming language $\PiPlusLang$.
\end{example}

\review{We characterise the path space of univalent subuniverses.}

\begin{proposition}
  If $T$ is an $n$-type, $\BAut[T]$ is an $(n+1)$-type.
\end{proposition}

\begin{proof}
  We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
  univalent subuniverse, we have $(X \id Y) \eqv (\fst(X) \eqv \fst(Y))$. Note that being an $n$-type is a proposition.
  Since $T$ is an $n$-type, and $\fst(X)$ and $\fst(Y)$ are merely equal to $T$, they're also $n$-types. It follows that
  $\fst(X) \eqv \fst(Y)$ is an $n$-type, and hence $X \id Y$ is an $n$-type.
\end{proof}

\begin{proposition}
  For any $T : \UU$, $\BAut[T]$ is 0-connected.
\end{proposition}

\begin{corollary}
  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid, or has h-level 3.
\end{corollary}

\begin{lemma}~\label{lem:loop-deloop}
  \[
    \loopspace[\BAut[T],T_{0}] \eqv \Aut[T]
  \]
\end{lemma}

\begin{proof}
  Since $\BAut[T]$ is a univalent universe, it follows that
  \[
    (T_{0} \id_{\BAut[T]} T_{0}) \eqv (\fst(T_{0}) \eqv \fst(T_{0})) \equiv (T \eqv T) \equiv \Aut[T].
  \]
\end{proof}

\begin{corollary}
  For every $n:\Nat$,
  \[
    \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]]
  \]
\end{corollary}
